Deriving Multi-Headed Planar Dependency Parses from Link Grammar Parses

### 2015-08-20 doc/slides/hong+eisner.TLT13.slides.tex
updated slides

\begin{document}
\title{Deriving Multi-Headed Planar Dependency Parses from Link Grammar Parses}
\author{Juneki Hong, Jason Eisner}
\date{\today}
\frame{\titlepage} %\NoteJH{Include pictures of us?}

\section{Introduction}
\frame{\frametitle{Introduction}
\begin{itemize}
\item This talk is about automatically converting from one annotation style to another.
\item The conversion could be hard, where information is fragmented or missing.

%This is often useful.  You may want to do it yourself.  
\item We use a general technique, Integer Linear Programming to recover this missing information.
\end{itemize}
}

@@

\section{Motivation}
%\subsection{Multi-headed}
\frame{\frametitle{Why Multi-headedness?}
  Multi-headedness Can Capture Additional Linguistic Phenomenon
  \begin{itemize}
    \item Control
    \item Relativization
    \item Conjunction
  \end{itemize}
}

\subsection{Control}
\frame{\frametitle{Control}
  \begin{figure}
    \begin{dependency}
      \begin{deptext}
        Jill \& likes \& to \& skip \\
      \end{deptext}
      \deproot[edge above, thick, hide label, edge unit distance = 1.5ex]{2}{}
      \depedge[edge above, thick, hide label]{2}{1}{}
      \depedge[edge below, ultra thick, hide label, edge style = {purple}, edge unit distance = 1.2ex]{4}{1}{}
      \depedge[edge above, thick, hide label]{2}{3}{}
      \depedge[edge above, thick, hide label]{3}{4}{}
    \end{dependency}
    \caption*{Jill is the subject of two verbs}
  \end{figure}

@@

\subsection{Relativization}
\frame{\frametitle{Relativization}
  \begin{figure}
    \begin{dependency}
      \begin{deptext}
        The \& boy \& that \& Jill \& skipped \& with \& fell \& down \\
      \end{deptext}
      \deproot[edge above, thick, hide label, edge unit distance = 2ex]{7}{}
      \depedge[edge above, thick, hide label]{2}{1}{}
      \depedge[edge above, thick, hide label]{2}{3}{}
      \depedge[edge above, thick, hide label]{3}{5}{}
      \depedge[edge above, thick, hide label]{5}{4}{}
      \depedge[edge above, thick, hide label]{5}{6}{}
      \depedge[edge above, thick, hide label, edge unit distance = 1.8ex]{7}{2}{}
      \depedge[edge above, thick, hide label]{7}{8}{}
      \depedge[edge below, ultra thick, hide label, edge style = {purple}, edge unit distance = 1.5ex]{6}{2}{}
    \end{dependency}
    \caption*{The boy is the object of \textit{with} as well as the subject of \textit{fell}.}
  \end{figure}
}

\subsection{Conjunction}
\frame{\frametitle{Conjunction}
  \begin{figure}
    \begin{dependency}
      \begin{deptext}
        Jack \& and \& Jill \& went \& up \& the \& hill \\
      \end{deptext}
      \deproot[edge above, thick, hide label, edge unit distance = 2ex]{4}{}
      \depedge[edge above, thick, hide label]{2}{1}{}
      \depedge[edge above, thick, hide label]{2}{3}{}
      \depedge[edge above, thick, hide label]{4}{2}{}
      \depedge[edge above, thick, hide label]{4}{5}{}
      \depedge[edge above, thick, hide label]{5}{7}{}
      \depedge[edge above, thick, hide label]{7}{6}{}

@@

@@

\frame{\frametitle{Link Grammars: Converting into a Directed Acyclic Graph}
  \begin{figure}
    \begin{dependency}
	  \begin{deptext}
	    - \& n-u \& v \& e \& e \& v \& v-d \& r \& n-u \& - \\
	    the \& matter \& may \& never \& even \& be \& tried \& in \& court \& . \\
	  \end{deptext}
	  \depedge[edge above, thick, edge style = {red}]{2}{3}{S}
	  \depedge[edge above, thick, edge style = {red}]{3}{6}{I}
	  \depedge[edge above, thick, edge style = {red}]{2}{1}{D}
	  \depedge[edge above, thick, edge style = {red}]{7}{8}{MV}
	  \depedge[edge above, thick, edge style = {red}]{6}{7}{P}
	  \depedge[edge above, thick, edge style = {red}]{8}{9}{J}
	  \deproot[edge above, thick, edge style = {red}]{2}{W}
	  \deproot[edge above, thick, edge style = {red}]{7}{WV}
	  \deproot[edge above, thick, edge style = {red}]{10}{X}
	  \depedge[edge above, thick, edge style = {red}]{6}{4}{E}
	  \depedge[edge above, thick, edge style = {red}]{6}{5}{E}
    \end{dependency}

    \caption*{Directionalize the edges}
  \end{figure} 
}

@@

\subsection{Corpus Building}
\frame{\frametitle{Corpus Building Strategy}
\begin{itemize}
  \item We start with a corpus of link grammar parses.
  \item We try to directionalize this corpus, choosing a direction for every link in every parse.
  \item We use an Integer Linear Program to find consistent directions for every link
\end{itemize}
}

\subsection{ILP Model}
\frame{\frametitle{Integer Linear Programming Model}
Encoded Constraints:
\begin{itemize}
  \item Connectedness\uncover<2->{: Every word has a link attaching it}
  \item Acyclicity\uncover<2->{: No cycles!}
  \item Consistency of Directionalized Links\uncover<2->{:

    Similar links oriented the same way}
\end{itemize}
}

@@

Variables: 
\begin{itemize}
\item[] $x_{ij}, x_{ji} \in \mathbb{Z} \geq 0$: orientation of each link
\item[] $x_{ij} + x_{ji} = 1$
\end{itemize}
\uncover<2->{\begin{center}
    A link can only be oriented left or right
\end{center}}

@@

\frame{\frametitle{Connectedness, Acyclicity}
\begin{itemize}
  \item[] Connectedness
    \begin{align}
      \sum_u x_{uv} & \geq 1 
    \end{align}
    \uncover<2->{\begin{center}
        A word has at least 1 parent
    \end{center}}

  \item[] Acyclicity
    \begin{itemize}
      \item[] Given that node $u$ is the parent of $v$
      \item[] $n_v$: length of the sentence containing node $v$
      \item[] $d_v \in [0, n_v]$: depth of the node from the root of the sentence
    \end{itemize}

    \begin{align}
      (\forall_u)\; d_v + (1 + n_v) \cdot (1 - x_{uv}) & \geq 1+d_u
    \end{align}
    \uncover<2->{\begin{center}
        The depth of a child is greater than the depth of the parent
    \end{center}}
\end{itemize}

\frame{\frametitle{Consistency of Directionalized Links \onslide<2->{with Slack}}
\begin{itemize}
  \item[] Consistency of Directionalized Links
    \begin{itemize}
      \item[] $r_L, \ell_L \in \{0,1\}$: whether links with label $L$ allowed left/right
    \end{itemize}

@@

    \begin{align}\label{eqn:obj}
      \min \left( \sum_L r_L + \ell_L \right) \onslide<2->{\cdot \frac{N_L}{4} + \sum_{ij}s_{ij}}
    \end{align}

    \onslide<2->{\begin{itemize}
      \item[] $s_{ij} \in \mathbb{R} \geq 0$: slack variable 
      \item[] $N_L$: Number of link tokens with label $L$
    \end{itemize}

    \begin{itemize}
      \item[] Slack allows a few links with label $L$ in disallowed directions
    \end{itemize}
    }

@@

  \begin{itemize}
  \item We worried that the recovered direction mapping might be unstable and sensitive to the input corpus. 
  \item We compared the results of increasing runs of sentences.
  \end{itemize}

@@

@@

\section{Conclusions}
\frame{\frametitle{Conclusions}
  \begin{itemize}
  \item Link Grammar parses can be oriented into connected DAGs
  \item new corpora for building multi-headed dependency parsers
  \item ILP can be used to build or annotate corpora. 
  \end{itemize}
}

>>>>

\begin{document}
\title{Deriving Multi-Headed Planar Dependency Parses from Link Grammar Parses}
\author{Juneki Hong and Jason Eisner}
\date{}%\today}
\frame{\titlepage} %\NoteJH{Include pictures of us?}

\section{Introduction}
\frame{\frametitle{Introduction}
\begin{itemize}
\item This talk is about converting from one annotation style to another.
\item The conversion could be hard, where information is fragmented, missing, or ambiguous.

%This is often useful.  You may want to do it yourself.  
\item We use a general technique, Integer Linear Programming to help us do this conversion.
\end{itemize}
}

@@

\section{Motivation, Overview}
%\subsection{Multi-headed}
\frame{\frametitle{Why Multi-headedness?}
  Multi-headedness Can Capture Additional Linguistic Phenomenon
  \begin{itemize}
    \item Control
    \item Relativization
    \item Conjunction
  \end{itemize}
}

\subsection{Control, Relativization, Conjunction}
\frame{\frametitle{Control}
  \begin{figure}
    \begin{dependency}
      \begin{deptext}
        Jill \& likes \& to \& skip \\
      \end{deptext}
      \deproot[edge above, thick, hide label, edge unit distance = 1.5ex]{2}{}
      \depedge[edge above, thick, hide label]{2}{1}{}
      \depedge[edge below, ultra thick, hide label, edge style = {purple}, edge unit distance = 1.2ex]{4}{1}{}
      \depedge[edge above, thick, hide label]{2}{3}{}
      \depedge[edge above, thick, hide label]{3}{4}{}
    \end{dependency}
    \caption*{Jill is the subject of two verbs}
  \end{figure}

@@

%\subsection{Relativization}
\frame{\frametitle{Relativization}
  \begin{figure}
    \begin{dependency}
      \begin{deptext}
        The \& boy \& that \& Jill \& skipped \& with \& fell \& down \\
      \end{deptext}
      \deproot[edge above, thick, hide label, edge unit distance = 2ex]{7}{}
      \depedge[edge above, thick, hide label]{2}{1}{}
      \depedge[edge above, thick, hide label]{2}{3}{}
      \depedge[edge above, thick, hide label]{3}{5}{}
      \depedge[edge above, thick, hide label]{5}{4}{}
      \depedge[edge above, thick, hide label]{5}{6}{}
      \depedge[edge above, thick, hide label, edge unit distance = 1.8ex]{7}{2}{}
      \depedge[edge above, thick, hide label]{7}{8}{}
      \depedge[edge below, ultra thick, hide label, edge style = {purple}, edge unit distance = 1.5ex]{6}{2}{}
    \end{dependency}
    \caption*{The boy is the object of \textit{with} as well as the subject of \textit{fell}.}
  \end{figure}
}

%\subsection{Conjunction}
\frame{\frametitle{Conjunction}
  \begin{figure}
    \begin{dependency}
      \begin{deptext}
        Jack \& and \& Jill \& went \& up \& the \& hill \\
      \end{deptext}
      \deproot[edge above, thick, hide label, edge unit distance = 2ex]{4}{}
      \depedge[edge above, thick, hide label]{2}{1}{}
      \depedge[edge above, thick, hide label]{2}{3}{}
      \depedge[edge above, thick, hide label]{4}{2}{}
      \depedge[edge above, thick, hide label]{4}{5}{}
      \depedge[edge above, thick, hide label]{5}{7}{}
      \depedge[edge above, thick, hide label]{7}{6}{}

@@

\subsection{Corpus Building}
\frame{\frametitle{Corpus Building Strategy}
\begin{itemize}
\item We start with some sentences and parse them with LG Parser
\item We take the undirected parses and try to directionalize them.
\item We use an ILP to assign consistent directions for each link type.
\end{itemize}

\begin{figure}
  \includegraphics[width=\linewidth]{images/corpusbuilding}
\end{figure}
}

@@

%\frame{\frametitle{Link Grammars: Converting into a Directed Acyclic Graph}
%  \begin{figure}
%    \begin{dependency}
%	  \begin{deptext}
%	    - \& n-u \& v \& e \& e \& v \& v-d \& r \& n-u \& - \\
%	    the \& matter \& may \& never \& even \& be \& tried \& in \& court \& . \\
%	  \end{deptext}
%	  \depedge[edge above, thick, edge style = {red}]{2}{3}{S}
%	  \depedge[edge above, thick, edge style = {red}]{3}{6}{I}
%	  \depedge[edge above, thick, edge style = {red}]{2}{1}{D}
%	  \depedge[edge above, thick, edge style = {red}]{7}{8}{MV}
%	  \depedge[edge above, thick, edge style = {red}]{6}{7}{P}
%	  \depedge[edge above, thick, edge style = {red}]{8}{9}{J}
%	  \deproot[edge above, thick, edge style = {red}]{2}{W}
%	  \deproot[edge above, thick, edge style = {red}]{7}{WV}
%	  \deproot[edge above, thick, edge style = {red}]{10}{X}
%	  \depedge[edge above, thick, edge style = {red}]{6}{4}{E}
%	  \depedge[edge above, thick, edge style = {red}]{6}{5}{E}
%    \end{dependency}
%    
%    \caption*{Directionalize the edges}
%  \end{figure} 
%}

@@

\subsection{ILP Model}
\frame{\frametitle{Integer Linear Programming Model}
Encoded Constraints:
\begin{itemize}
  \item Acyclicity\uncover<2->{: (\textbf{No cycles!})}
  \item Connectedness\uncover<3->{: (\textbf{Every word is reachable from a root})}
  \item Consistency of Directionalized Links\uncover<4->{:

    (\textbf{Similar links oriented the same way})}
\end{itemize}

\begin{figure}
  \begin{subfigure}[b]{0.3\textwidth}
    \uncover<2->{\includegraphics[width=\linewidth]{images/acyclic2}}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\textwidth}
    \uncover<3->{\includegraphics[width=\linewidth]{images/connectedness}}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\textwidth}
    \uncover<4->{\includegraphics[width=\linewidth]{images/consistency}}
  \end{subfigure}

\end{figure}

@@

Variables: 
\begin{itemize}
\item[] $x_{ij}, x_{ji} \in \mathbb{Z} \geq 0$: orientation of each link
\item[] $x_{ij} + x_{ji} = 1$
\end{itemize}
\uncover<2->{\begin{center}
    \textbf{An individual link token can either be oriented left or oriented right}
\end{center}}

@@

\frame{\frametitle{Acyclicity, Connectedness}
\begin{itemize}
  \item[] Acyclicity
    \begin{itemize}
      \item[] Given that node $u$ is the parent of $v$
      \item[] $n_v$: length of the sentence containing node $v$
      \item[] $d_v \in [0, n_v]$: depth of the node from the root of the sentence
    \end{itemize}

    \begin{align}
      (\forall_u)\; d_v + (1 + n_v) \cdot (1 - x_{uv}) & \geq 1+d_u
    \end{align}
    \uncover<2->{\begin{center}
        \textbf{The depth of a child is greater than the depth of the parent}
    \end{center}}

  \item[] Connectedness
    \begin{align}
      \sum_u x_{uv} & \geq 1 
    \end{align}
    \uncover<3->{\begin{center}
        \textbf{A word has at least 1 parent}
    \end{center}}
\end{itemize}
}

\frame{\frametitle{Consistency of Directionalized Links \onslide<2->{with Slack}}
\begin{itemize}
  \item[] Consistency of Directionalized Links
    \begin{itemize}
      \item[] $r_L, \ell_L \in \{0,1\}$: whether all links with label $L$ allowed left/right
    \end{itemize}

@@

    Objective Function:
    \begin{align}\label{eqn:obj}
      \min \left( \sum_L r_L + \ell_L \right) \onslide<2->{\cdot \frac{N_L}{4} + \sum_{ij}s_{ij}}
    \end{align}

    \onslide<2->{\begin{itemize}
      \item[] $s_{ij} \in \mathbb{R} \geq 0$: slack variable 
      \item[] $N_L$: Number of link tokens with label $L$
    \end{itemize}
    }
    \onslide<3->{
    \begin{itemize}
      \item[] \textbf{Slack allows a few links with label $L$ in disallowed directions}
    \end{itemize}
    }

@@

  \begin{itemize}
  \item We were worried that the recovered direction mapping might be unstable and sensitive to the input corpus. 
  \item We compared the results of increasing runs of sentences.
  \end{itemize}

@@

\frame{\frametitle{ILP Results: Top 25 Most Occurring Labels}
  \begin{figure}
    \centering
    \input{link_table_filtered_highlighted.tex}
  \end{figure}
}

\frame{\frametitle{ILP Results: Top 25 Most Occurring Labels}
  \begin{figure}
    \centering
    \input{link_table_filtered_highlighted_explanation.tex}
  \end{figure}
}

@@

\section{Conclusions}
\frame{\frametitle{Conclusions}
  \begin{itemize}
  \item Link Grammar parses can be oriented into connected DAGs
  \item A new corpus available for building multi-headed dependency parsers
  \item ILP can be used to help annotate incomplete or missing data in corpora. 
  \end{itemize}
}
